Developing decarbonisation pathways for Irish homes in changing TIMES

### 2021-08-27 manuscript.tex
Done for 27 Aug 2021

The models described in this section include a range of national ESOMs and bottom-up residential models with varied objectives. The complexity of models has increased significantly in the last 20 years, in Ireland, the models started out in Microsoft Excel and switched to Microsoft SQL, the higher computational power allows for deeper analyses. The growing BER database is provides detailed open sourced data for input to Irish models. 
%The internal temperature is becoming more frequently analysed in Irish residential studies, a recurring theme is the wide range of internal temperatures which Dineen et al. \cite{Dineen2015ImprovedSource} found this assumption significantly changes the final residential energy consumption. 
\subsection{Case study for Ireland}
\label{Intro:IreRSD}

@@

Space and water heating are the main energy services, accounting for 80\% of the energy used in residential dwellings \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and typically have a lifetime of at least 15 years, consequently leading to a slow turnover of technology in the residential sector. Apart from space and water heating, all EU member states are required under commission regulation (EU) 431/2014 to supply energy consumption data on space cooling, cooking, lighting \& appliances, and others (mainly outdoor appliances), the annual trend of these energy services is shown in Figure \ref{fig:IrelandServiceDemands}. These energy services were also used in TIM, however there is no space cooling data available in Ireland and lighting \& appliances were separated while others was combined with appliances.

@@

The Irish government's long-term overarching strategy aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, [as] a top priority"}\cite{Ireland2018ProjectFramework}. The densification of Ireland's main cities will provide opportunities to improve the efficiency of heating, electricity, and transport within and between cities. 
\par The Marginal Abatement Cost Curve (MACC) produced in CAP2019, shows electricity \& transport are the most cost-effective sectors for energy decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}. In Ireland, the decarbonisation of electricity is underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2020Energy-RelatedIreland.}. 
Ireland uses both the \textit{carrot} (subsidy) and \textit{stick} (tax) approach to direct private investment towards low carbon or renewable energy. A €15/ $tCO_2$ carbon tax was introduced in 2009 to reduce investment in fossil fuels, and the increasing carbon tax rate has resulted in tax receipts growing every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}. Ireland's current carbon tax of €33.50/ $tCO_2$ is set to rise to €100/ $tCO_2$ in 2030. The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentivise renewable electricity investment to achieve legally binding renewable targets under 2009/28/EC. The Renewable Electricity Support Scheme (RESS) replaced REFIT in 2020. RESS supports renewable electricity projects to help Ireland achieve 2030 climate targets, especially the 70\% renewable electricity target.

\par As the grant provider, SEAI plays a key role in the rollout of retrofitting and heat pumps. The retrofitting and heat pump grants do still require large upfront costs from the consumer, which may hinder the ambitious targets as this wont be a viable option for many households, green loans are currently available to bridge this gap. % are therefore the preferred heating technology despite the high capital costs. CAP2019 planned to retrofit 500,000 existing homes to at least a  ‘B2’ energy rating by 2030, which highlights the energy efficiency first approach in the residential sector. 
\par Gas Network Ireland (GNI) have set a 18\% biomethane target for 2030 \cite{Ervia2019VisionIreland}, where CAP2019 only set a 3\% biomethane target for 2030 \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}. %GNI anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply, which provided 61\% of gas in 2018 ) will be less than 30\% of 2018 levels \cite{OCleirigh2020EnergySEAI}. 
A major barrier to increasing biomethane from an agricultural source in the gas network is the cost, it was the most expensive option in the MACC produced in CAP2019.
\par Ireland has one of the lowest shares of DH in Europe at less than 1\%  \cite{Gartland2016AIreland}. %
%Advancement in DH control and awareness of the technology will strengthen the case for DH networks to play a part in decarbonising the residential sector.
CAP2019 plans to have approx. 0.4\% of heating provided by DH in 2030. The Irish District Energy Association (IrDEA) suggest 57\% of heat can be sourced from DH, if supporting policy and regulation was in place. CAP2019 sets out a need to develop a policy framework for the development of DH in Ireland 
%DH will likely play a role, however, Gartland points out a range of organisational, technical, regulatory, and economic barriers to DH growth in Ireland \cite{Gartland2016AIreland}.

@@

The distinctive Irish residential sector provides an interesting case study and new Irish climate targets now means previous Irish models which relied on superseded climate targets are out-dated, but not redundant. A new iteratively built transparent ESOM with wide stakeholder involvement, which is reflective of current climate policy is needed to robustly explore optimal decarbonisation pathways and technical limits. The new model will provide opportunity for deeper analysis of the energy system sectors, with accessible soft-linking capabilities. The model development process can be applied to any national or regional ESOM.  
No such ESOM with a focus on Irish residential sector exists, the strength of an ESOM is to account for sectoral interactions which bottom-up simulation models cannot provide. Structural improvements and archetypes in the model aligns with existing retrofitting and electrification strategies. Other model innovations include filtering the BER database, other EU countries with publicly available BER database under directive (EU) 2018/844 can apply similar methods, while TIM also considers internal temperatures shifts and new energy service demands were calculated. TIM will be a critical tool in exploring  
\textit{how changes to future technology developments affect the optimal residential heating decarbonisation pathway in Ireland? }

@@

%``How should Ireland set GHG targets in the residential sector?”

@@

\par  The expertise among energy policy researchers in Ireland combined with the global network of Energy Technology Systems Analysis Program (ETSAP) modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways. Previous versions of the Irish TIMES model were extracted from the Pan European TIMES model and then updated with specific national data and assumptions. The previous Irish TIMES models were used to provide informative reports on Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}.  
The introduction of new climate policies and targets combined with technological advancements, means there is a need for a new ESOM to be built to explore Ireland's new climate targets and sectoral five-year carbon budgets, the new ``TIMES Ireland Model" (TIM) was built in 2021 to address this need \textbf{(SOURCE)}.
\par TIMES is a bottom-up optimisation energy-environment model with perfect foresight which is used to analyse various levels of spatial, temporal and sectoral resolution. Exogenous technical, physical, and regulatory constraints can be applied and technologies have specified fuel types, efficiency, lifetime, availability, environmental characteristics, and fixed, variable and operation costs. TIM is set up with regional options, either one national region or 26 county regions, only the annual results were explored in this study. TIMES uses a linear programming optimizer matrix in General Algebraic Modeling System (GAMS), to compute all calculations.

@@

TIM was developed completely online during national lockdowns, with no face to face meetings, the model relied on git-centred model development process with different developers leading different sectors. Weekly meetings were combined with intensive daily ``sprint" meetings to keep model milestones on track. The transparency of TIM improved the iterative development process, external energy experts and other stakeholders contributed using web-based dashboards \footnote{https://tim-review1.netlify.app/results/}. The schematic in Figure \ref{fig:TIMOverviewRES} shows the main sectors of the model, represented in the demand-side modules and how they link with the supply-side modules, to create a full picture of the national energy system \textbf{SOURCE}.

@@

@@

The residential sector was developed with awareness of the ongoing national retrofitting and electrification strategies and investments. The residential energy services were developed based upon regulation (EU) 431/2014, the BER database and EU databases. Figure \ref{fig:RES} illustrates the flow of energy from fuel to demand in the residential sector.

@@

The BER database provides the base year bottom-up energy consumption data, this is scaled up to represent the entire stock and cross-checked with the top-down energy consumption from the 2018 SEAI Energy Balance. To compare from the two databases, fuels were were combined and condensed down to 13 fuels in TIM. Table \ref{AppTable:Fuels} shows an overview of the fuels combined from the different data sources. The full list of space and water heating technologies considered in the model can be found in Table \ref{AppHeatingTech}.

\par In TIM, fuels such as ``coal", represent a combination of fuels from the SEAI Energy Balance and BER database. Gasoil which has similar properties to kerosene, is accounted for in the SEAI Energy Balance,  but there is no evidence to support keeping gasoil and kerosene separate would provide further insight into residential decarbonisation, therefore these fuels are combined in TIM. However, disaggregating ambient heat could potentially provide insights on residential decarbonisation, so the model has separated space ambient heat and water ambient heat. The solid multi-fuel from the BER database represents an open fire or stove which is heated by either coal, peat, or wood. Through examining the SEAI Energy Balance and BER database, the fuel share for coal, peat, and wood in solid multi-fuel was set to 19.4\%, 74.4\% and 6.2\% respectively. 
\par There are two types of energy services represented in Fig. \ref{fig:RES} - Non-Archetype and Archetype. Essentially Non-archetype energy services have a lack of data and could not be accurately assigned to an archetype. The six non-archetype energy services are not dependant on the type of dwelling in the model and the base year quantity is the total national consumption of that energy service. Non-Archetype data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA database and the top-down data from the SEAI Energy Balance, more details on non-archetype services are found in Section \ref{Meth:Data}. The four archetype energy services are dependent on  building type and data is sourced from the BER database, which is discussed in Section \ref{Meth:Data}. The internal temperature adjustment for space heating is also detailed in Section \ref{Meth:Data}. The demand for archetype and non-archetype energy services is based on the growth of residential dwellings, which is detailed in section \ref{Meth:Data}.

\subsubsection{Data}
\label{Meth:Data}
%
The dwelling stock data is obtained from the SEAI BER Public database \footnote{https://ndber.seai.ie/BERResearchTool/ber/search.aspx}, which is firstly filtered, then scaled up using Central Statistics Office (CSO) data. 
The SEAI BER public database is Ireland's national Energy Performance Certificate (EPC) register,  which complies with the Energy Performance of Buildings Directive 2018/844/EU (EPBD). The BERs are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$). 
\par Over half of Irish dwellings are captured in the BER database and every BER displays over 200 energy related variables. The DEAP model which is used to assign BER labels,  is outlined by \citet{Dineen2015ImprovedSource} in Section \ref{Intro:ESOM}. In terms of space heating, DEAP assumes that each dwelling is single zoned, the heating times are 7am to 9am and 5pm to 11pm ( 56 hours / week ) during the heating season, which is from October to May inclusive and assumes monthly external temperatures, the average external temperature for the heating season is 7.6$^{\circ}$C. One assumption, the internal temperature assumption, results in a large error, where space and water heating are overestimated in DEAP. Space heating consumption is reduced by 26\% after internal temperature adjustments were applied in TIM, this brought the final energy consumption in line with the top-down SEAI Energy Balance. DEAP assumes that living areas ( living room, sitting room, and possibly kitchen) in all dwellings are heated to 21$^{\circ}$C and non-living areas ( bedroom, bathroom, hallways ) are heated to 18$^{\circ}$C. Through the analysis of previous residential modelling in Ireland this is not the case, as the internal temperatures are often much lower, Table \ref{Table:InternalTemps} outlines the residential internal temperature assumptions applied in TIM.

@@

The ArDEM-SQL, which was described in section \ref{Intro:ESOMIreland} is a simulation model built on space heating calculation EN 13790. ArDEM-SQL alters the primary energy consumption of the BER database by changing the internal temperatures assumed by DEAP in Table \ref{Table:InternalTemps}.

Other concerns for uncertainties in the database, which are not related to DEAP calculation are statistically anomalous spikes in the data, caused by applying default values \cite{Ahern2020ADatabases} and different companies generating variations of ±20\% in the assessment of energy consumption \cite{Harsman2016OnCertificates}. While the BER database provides high granularity data, it must be filtered to improve the representation of the dwelling stock data. The filters applied to the raw BER database are outlined in Table \ref{AppBERFilters}.

The Census 2016 data \cite{CentralStatisticsOffice2016CensusIreland} from the CSO is also used in calculating the dwelling stock. Dataset E1005 provides a split of seven building archetypes per county, which are aggregated to form three building archetypes. The total number of dwellings per archetype per county in 2016 is used to scale up the BER database. The gap in total dwellings between the census year in 2016 and the model base year in 2018 was initially ignored due to lack of data. However, later it was found that data on new dwelling electrical connections per archetype and per county are available from CSO dataset NDQ06, this would have filled the gap, the first iteration of the dwelling stock in TIM has a 2\% error. 
\par Once the BER database and CSO database tables have been filled to include dwellings per archetype per county, the next step was to scale up the BER database. This was done by first finding the ratio of the building archetype per county between the BER database and CSO data, as shown in Equation \ref{Eq:BERtoCSO1}.

@@

Where $TIM_{r,y,a}$ is the total dwellings in TIM by BER rating (r), year of construction (y), and archetype (a). $CSO_{y,a}$ is the number of dwellings by year of construction and archetype in the CSO dataset, the CSO dataset does not contain BER rating types. $BER_{r,y,a}$ is the number of BER ratings by year of construction and archetype from the BER database and dividing this value by $BER_{y,a}$ results in a probability of a certain age type falling within a BER rating, and the final part of the equation $\frac{TIM_{a}} {BER_{a}}$ compensates for the non-stated archetypes in the CSO data, $TIM_a$ was already calculated in equation \ref{Eq:BERtoCSO2}. The results of each of the correlations are in  Table \ref{AppBERCor},\ref{AttBERCor} \& \ref{DetBERCor} and Table \ref{Table:FinalDwellingStock} outlines the final dwelling stock used in TIM.

@@

The costs applied in the residential sector, which includes space and water heating technologies, retrofitting, cooking, lighting, pumps, fans, space cooling, and other electrical appliances. 
The fuel costs are not applied in the residential sector, there are applied in the supply sector, and the data is sourced from World Energy Outlook reports from 2019 \cite{/content/publication/caf32f3b-en} and 2020 \cite{/content/publication/557a761b-en}. 
The majority of the space and water heating cost data is obtained from the
\textit{``Techno-economic projections until 2050 for smaller heating and cooling technologies in the residential and tertiary sectors in the EU”} report \cite{JointResearchCentre2017Techno-economicEU}. The cost data used in TIM was based on the north EU region (Denmark) and where data was missing, the central EU region (Germany) cost data was used. The high granularity of data allows for the cost of a technology to be disaggregated between apartments and non-apartments, costs and efficiencies are projected every five years from 2020 to 2050. The lifetime and emissions/pollutants are also included, while the cost data is further disaggregated between capital expenditure ( equipment, installation \& additional capital costs ) and Operational \& Maintenance (O\&M) costs.  
\par TIM has two types of retrofitting options per archetype,  a shallow retrofit is when the primary energy ($kWh/m^2/yr$) is reduced by $ \leq 35\%$, and deep retrofit is where the primary energy is reduced by $>35\%$, while a reduction of $ \ >75\%$ is considered infeasible. The difference between the pre-retrofit and post-retrofit primary energy determines the type of retrofit, as shown in Table \ref{AppDepthofRetrofit}.

The retrofit cost data was sourced from Ali et. al \cite{Ali2020ABuildings} who derived costs from SEAI's Better Energy Homes (BEH) and Better Energy Warmer Homes (BEWH) datasets, combined this is Ireland's largest retrofitting dataset ( over 285,000 retrofits ). To provide more robust retrofit costs, data is also used from the Typology Approach for Building Stock Energy Assessment (TABULA) report \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland} which has more detailed cost data with two stages of retrofit (standard and advanced) for a sample of 31 dwellings. The formulas used to find the average cost of a retrofit per type of retrofit and archetype are shown in Equations \ref{Eq:RetrofitCost1} and \ref{Eq:RetrofitCost2}.

@@

Where $ {\text{\euro}}{}_{rf-rt,a}$ is the retrofit cost from pre-retrofit BER rating (rf) to post-retrofit rating (rt) for archetype (a). $Ali_{r,t}$ is the cost data from  \citet{Ali2020ABuildings}, it contains all BER ratings and retrofit types but does not include an archetype, so an archetype factor (AF) is introduced, 0.9 for apartments, 1 for attached, and 1.1 for detached dwellings.  $TABULA_{rf-rt,a}$ is the average retrofit cost from pre-retrofit BER rating to post-retrofit BER rating for an archetype, data is sourced from TABULA \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland}.  However with only 62 samples in total not all retrofits are included. In this case, TABULA is ignored and \citet{Ali2020ABuildings} is the only source of retrofit costs. The retrofit costs are cross-checked with the residential cost-optimal report \cite{AECOM2020ReportBuildings}. 
\par To simplify the model, the cost of retrofit per BER rating is not used in TIM. Instead, an average weighted cost per archetype and per retrofit is calculated.

>>>>

The models described in this section include a range of national ESOMs and bottom-up residential models with varied objectives. The complexity of models has increased significantly in the last 20 years, in Ireland, the models started out in Microsoft Excel and switched to Microsoft SQL, the higher computational power allows for deeper analyses. The growing BER database is provides detailed open sourced data for input to Irish models. 
%The internal temperature is becoming more frequently analysed in Irish residential studies, a recurring theme is the wide range of internal temperatures which Dineen et al. \cite{Dineen2015ImprovedSource} found this assumption significantly changes the final residential energy consumption.

\subsection{Case study for Ireland}
\label{Intro:IreRSD}

@@

Space and water heating are the main energy services, accounting for 80\% of the energy used in residential dwellings \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and typical heating systems have a lifetime of at least 15 years, consequently leading to a slow turnover of technology in the residential sector. Apart from space and water heating, all EU member states are required under commission regulation (EU) 431/2014 to supply energy consumption data on space cooling, cooking, lighting \& appliances, and others (mainly outdoor appliances), the annual trend of these energy services is shown in Fig. \ref{fig:IrelandServiceDemands}.

@@

The Irish government's long-term overarching strategy aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, [as] a top priority"}\cite{Ireland2018ProjectFramework}. The densification of Ireland's main cities will provide opportunities to improve the efficiency of heating, electricity, and transport within and between cities. The Marginal Abatement Cost Curve (MACC) produced in CAP2019, shows electricity \& transport are the most cost-effective sectors for energy decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}. In Ireland, the decarbonisation of electricity is underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2020Energy-RelatedIreland.}. 
Ireland uses both the \textit{carrot} (subsidy) and \textit{stick} (tax) approach to direct private investment towards low carbon or renewable energy. A €15/ $tCO_2$ carbon tax was introduced in 2009 to reduce investment in fossil fuels, and the increasing carbon tax rate has resulted in tax receipts growing every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}. Ireland's current carbon tax of €33.50/ $tCO_2$ is set to rise to €100/ $tCO_2$ in 2030. The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentivise renewable electricity investment to achieve legally binding renewable targets under 2009/28/EC. The Renewable Electricity Support Scheme (RESS) replaced REFIT in 2020. RESS supports renewable electricity projects to help Ireland achieve 2030 climate targets, especially the 70\% renewable electricity target.

\par As the grant provider, SEAI plays a key role in the rollout of retrofitting and heat pumps. The retrofitting and heat pump grants do still require large upfront costs from the consumer, which may hinder the ambitious targets as this wont be a viable option for many households, green loans are currently available to bridge this gap and other options are being explored. % are therefore the preferred heating technology despite the high capital costs. CAP2019 planned to retrofit 500,000 existing homes to at least a  ‘B2’ energy rating by 2030, which highlights the energy efficiency first approach in the residential sector. 
Gas Network Ireland (GNI) have set a 18\% biomethane target for 2030 \cite{Ervia2019VisionIreland}, where CAP2019 only set a 3\% biomethane target for 2030 \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}. %GNI anticipates that by 2026 or 2027 the supply from Corrib ( Irelands only indigenous gas supply, which provided 61\% of gas in 2018 ) will be less than 30\% of 2018 levels \cite{OCleirigh2020EnergySEAI}. 
A major barrier to increasing biomethane from an agricultural source in the gas network is the cost, it was the most expensive option in the MACC produced in CAP2019.
Ireland has one of the lowest shares of DH in Europe at less than 1\%  \cite{Gartland2016AIreland}. %
%Advancement in DH control and awareness of the technology will strengthen the case for DH networks to play a part in decarbonising the residential sector.
CAP2019 plans to have 0.4\% of heating provided by DH in 2030. The Irish District Energy Association (IrDEA) suggest 57\% of heat can be sourced from DH, if supporting policy and regulation was in place. CAP2019 sets out a need to develop a policy framework for the development of DH in Ireland.
%DH will likely play a role, however, Gartland points out a range of organisational, technical, regulatory, and economic barriers to DH growth in Ireland \cite{Gartland2016AIreland}.

@@

The distinctive Irish residential sector provides an interesting case study and new Irish climate targets now means previous Irish models which relied on superseded climate targets are out-dated, but not redundant. A new iteratively built transparent ESOM with wide stakeholder involvement, which aligns with current climate policy is needed to robustly explore optimal decarbonisation pathways and technical limits. 
The new model will provide opportunity for deeper analysis of the energy system sectors, with accessible soft-linking capabilities. The model development process can be applied to any national or regional ESOM.  
No such ESOM with a focus on Irish residential sector exists, the strength of an ESOM is to account for sectoral interactions which bottom-up simulation models cannot provide. Structural improvements and archetypes in the model aligns with existing retrofitting and electrification strategies. Other model innovations include filtering the BER database, other EU countries with publicly available BER database under directive (EU) 2018/844 can apply similar methods, while TIM also considers internal temperatures shifts and new energy service demands were calculated. TIM can provide a robust methodology into \textit{``how Ireland should set GHG targets in the residential sector”} and TIM will be a critical tool in exploring \textit{how changes to future technology developments affect the optimal residential heating decarbonisation pathway in Ireland? }.

@@

@@

\par  The expertise among energy policy researchers in Ireland combined with the global network of Energy Technology Systems Analysis Program (ETSAP) modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways. Previous versions of the Irish TIMES model were extracted from the Pan European TIMES model and then updated with specific national data and assumptions, to provide informative reports on Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}. The previous Irish TIMES model was non-transparent and inflexible model, which was not purpose built for the Irish energy system. The introduction of new climate policies and targets combined with technological advancements pushed the need for a new model, however the need for a transparent, flexible and robust model, purpose built for the Irish energy system by Irish energy modellers was a key driver to develop a new model.The new ``TIMES Ireland Model" (TIM) was built in 2021 to address this need \textbf{(SOURCE)}.
\par TIMES is a bottom-up optimisation energy-environment model with perfect foresight which is used to analyse various levels of spatial, temporal and sectoral resolution. Exogenous technical, physical, and regulatory constraints can be applied and technologies have specified fuel types, efficiency, lifetime, availability, environmental characteristics, and fixed, variable and operation costs. TIMES uses a linear programming optimizer matrix in General Algebraic Modeling System (GAMS), to compute all calculations.

@@

TIM was developed completely online during national lockdowns, with no face to face meetings, the model relied on git-centred model development process with different developers leading different sectors. Weekly meetings were combined with intensive daily ``sprint" meetings to keep model milestones on track. The transparency of TIM improved the iterative development process, external energy experts and other stakeholders contributed using web-based dashboards\footnote{https://tim-review1.netlify.app/results/}. The schematic in Fig. \ref{fig:TIMOverviewRES} shows the main sectors of the model, represented in the demand-side modules and how they link with the supply-side modules, to create a full picture of the national energy system \textbf{SOURCE}.

@@

TIM is set up with regional options, either one national region or 26 county regions, and with different temporal resolution options, only the national annual results were explored in this study.

@@

The residential sector was developed with awareness of the ongoing national retrofitting and electrification strategies and investments. The residential energy services were developed based upon regulation (EU) 431/2014, shown in Fig. \ref{fig:IrelandServiceDemands}, however there is no space cooling data available in Ireland and lighting \& appliances are separated in TIM, while ``others" was combined with appliances. An overview of the flow of energy from fuels to demand, through the energy services is shown in Fig. \ref{fig:RES}.

@@

The BER database provides the base year bottom-up energy consumption data, this is scaled up to represent the entire stock and cross-checked with the top-down energy consumption from the 2018 SEAI Energy Balance. To compare the two databases, fuels were were merged and separated, with 13 fuels adequately covering the residential sector in TIM, Table \ref{AppTable:Fuels} summarizes the fuels from the different data sources. In TIM, fuels such as ``coal", represent a combination of fuels from the SEAI Energy Balance and BER database. Gasoil which has similar properties to kerosene, is accounted for in the SEAI Energy Balance,  but there is no evidence to support keeping gasoil and kerosene separate would provide further insight into residential decarbonisation, therefore these fuels are combined in TIM. However, disaggregating ambient heat could potentially provide insights on residential decarbonisation, so the model has separated space ambient heat and water ambient heat. The solid multi-fuel from the BER database represents an open fire or stove which is heated by either coal, peat, or wood. Through examining the SEAI Energy Balance and BER database, the fuel share for coal, peat, and wood in solid multi-fuel was set to 19.4\%, 74.4\% and 6.2\% respectively. 
\par There are two types of energy services represented in Fig. \ref{fig:RES} - Non-Archetype and Archetype. Essentially Non-archetype energy services have a lack of data and could not be accurately assigned to an archetype. The six non-archetype energy services are not dependant on the type of dwelling in the model and the base year quantity is the total national consumption of that energy service. Non-Archetype data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA database and the top-down data from the SEAI Energy Balance, more details on non-archetype services are found in Section \ref{Meth:Data}. The four archetype energy services and the internal temperature adjustment are dependent on  building type and data is sourced from the BER database, which is discussed in Section \ref{Meth:Data}. The demand for archetype and non-archetype energy services is based on the growth of residential dwellings, which is also detailed in Section \ref{Meth:Data}.

\subsubsection{Data}
\label{Meth:Data}
%
The dwelling stock data is obtained from the SEAI BER Public database\footnote{https://ndber.seai.ie/BERResearchTool/ber/search.aspx}, is filtered then scaled up using Central Statistics Office (CSO) data. 
The SEAI BER public database is Ireland's national Energy Performance Certificate (EPC) register,  which complies with the Energy Performance of Buildings Directive 2018/844/EU (EPBD). The BERs are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$). Over half of all Irish dwellings are captured in the BER database and every BER displays over 200 energy related variables. The DEAP model which is used to assign BER labels,  is outlined by \citet{Dineen2015ImprovedSource} in Section \ref{Intro:ESOM}. In terms of space heating, DEAP assumes that each dwelling is single zoned, the heating times are 7am to 9am and 5pm to 11pm ( 56 hours / week ) during the heating season, which is from October to May inclusive and assumes monthly external temperatures, the average external temperature assumed for the heating season is 7.6$^{\circ}$C. However, one assumption, the internal temperature assumption, results in a large unsupportable error, where space heating is overestimated by 26\%, this is a simliar finding to \citet{Dineen2015ImprovedSource} who found a 7.5\% change in energy consumption for every 1$^{\circ}$C shift in internal temperatures. After internal temperature adjustments in Table \ref{Table:InternalTemps} were applied in TIM, this brought the final energy consumption in line with the top-down SEAI Energy Balance. DEAP assumes that living areas ( living room, sitting room, and possibly kitchen) in all dwellings are heated to 21$^{\circ}$C and non-living areas ( bedroom, bathroom, hallways ) are heated to 18$^{\circ}$C. Through the analysis of previous residential modelling in Ireland internal temperatures are often much lower than 21$^{\circ}$C in living areas and 18$^{\circ}$C in non-living areas \cite{Dineen2015ImprovedSource,Healy2002FuelIreland,Clinch2003ValuingModel,Clinch2000HousingMortality}.

@@

The ArDEM-SQL, which was described in section \ref{Intro:ESOMIreland} is a simulation model built on space heating calculation EN 13790. ArDEM-SQL alters the primary energy consumption of the BER database by changing the internal temperatures assumed by DEAP in Table \ref{Table:InternalTemps}. Other concerns for uncertainties in the database, which are not related to DEAP calculation are statistically anomalous spikes in the data, caused by applying default values \cite{Ahern2020ADatabases} and different companies generating variations of ±20\% in the assessment of energy consumption \cite{Harsman2016OnCertificates}. While the BER database provides high granularity data, it must be filtered to improve the representation of the dwelling stock data. The filters applied to the raw BER database are outlined in Table \ref{AppBERFilters}. 
%PICKUPHERE
The Census 2016 data \cite{CentralStatisticsOffice2016CensusIreland} from the CSO is also used in calculating the dwelling stock, dataset E1005 provides a split of seven building archetypes per county, which are aggregated to form three building archetypes in TIM. The total number of dwellings per archetype per county in 2016 is used to scale up the BER database. The gap in total dwellings between the census year in 2016 and the model base year in 2018 was initially ignored due to lack of data. However, later it was found that data on new dwelling electrical connections per archetype and per county are available from CSO dataset NDQ06, this would have filled the gap, the first iteration of the dwelling stock in TIM does not account for the 2\% of existing dwellings build in 2017 and 2018. 
\par Once the BER database and CSO database tables have been filled to include dwellings per archetype per county, the next step was to scale up the BER database. This was done by first finding the ratio of the building archetype per county between the BER database and CSO data, as shown in Equation \ref{Eq:BERtoCSO1}.

@@

Where $TIM_{r,y,a}$ is the total dwellings in TIM by BER rating (r), year of construction (y), and archetype (a). $CSO_{y,a}$ is the number of dwellings by year of construction and archetype in the CSO dataset, no BER rating types exist in the CSO dataset. $BER_{r,y,a}$ is the number of BER ratings by year of construction and archetype from the BER database and dividing this value by $BER_{y,a}$ results in a probability of a certain age type falling within a BER rating, and the final part of the equation $\frac{TIM_{a}} {BER_{a}}$ compensates for the non-stated archetypes in the CSO data, $TIM_a$ was already calculated in Equation \ref{Eq:BERtoCSO2}. The results of each of the BER probablities are summarized for each archetype in  Table \ref{AppBERCor},\ref{AttBERCor} \& \ref{DetBERCor}, while Table \ref{Table:FinalDwellingStock} outlines the final dwelling stock used in TIM.

@@

The costs applied in the residential sector, includes space and water heating technologies, retrofitting, cooking, lighting, pumps, fans, space cooling, and other electrical appliances. 
The majority of the space and water heating cost data is obtained from the
\textit{``Techno-economic projections until 2050 for smaller heating and cooling technologies in the residential and tertiary sectors in the EU”} report \cite{JointResearchCentre2017Techno-economicEU}, the report has spatial cost data and the north EU region (Denmark) was mainly used \cite{JointResearchCentre2017Techno-economicEU}, this cost data is sourced from the continuously updated and expanding Danish Energy Agency's Technology Data for Individual Heat Production dataset and where data was missing, the central EU region (Germany) cost data was used \cite{JointResearchCentre2017Techno-economicEU}. The high granularity of data allows for the cost of a technology to be disaggregated between apartments and non-apartments, costs and efficiencies are projected every five years from 2020 to 2050. The lifetime and emissions/pollutants are also included, while the cost data is further disaggregated between capital expenditure ( equipment, installation \& additional capital costs ) and Operational \& Maintenance (O\&M) costs. 
The cost data for different electrical appliances is sourced from Topten EU database \cite{Michel2016EnergyREPORT} and SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}. Topten has a large database covering electrical appliances available in the European market, the data collected includes cost and efficiency of refrigeration, cloth washing, cloth drying, and dish washing, while some Topten reports have looked at sales and European market shares \cite{Michel2016EnergyREPORT}. Cooking data is sourced from Eurostat's ``Energy consumption in households by type and end-used"  \cite{Eurostat2021Disaggregatednrg_d_hhq} and cross-checked with SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}. Appliances not included in this list are simply under the ``Electrical Appliances" category in TIM. Electrical Appliances include television, multimedia, outdoor appliances, and computer technology and data is sourced from SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and Eurostat \cite{Eurostat2021Disaggregatednrg_d_hhq}. The fuel costs are not applied in the residential sector, there are applied in the supply sector, and the data is sourced from World Energy Outlook reports from 2019 \cite{/content/publication/caf32f3b-en} and 2020 \cite{/content/publication/557a761b-en}.  
\par TIM has two types of retrofitting options per archetype,  a shallow retrofit is when the primary energy ($kWh/m^2/yr$) is reduced by $ \leq 35\%$, and deep retrofit is where the primary energy is reduced by $>35\%$, while a reduction of $ \ >75\%$ is considered infeasible. The difference between the pre-retrofit and post-retrofit primary energy determines the type of retrofit, as shown in Table \ref{AppDepthofRetrofit}. The retrofit cost data was sourced from Ali et. al \cite{Ali2020ABuildings} who derived costs from SEAI's Better Energy Homes (BEH) and Better Energy Warmer Homes (BEWH) datasets, combined this is Ireland's largest retrofitting dataset ( over 285,000 retrofits ). To provide more robust retrofit costs, data is also used from the Typology Approach for Building Stock Energy Assessment (TABULA) report \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland} which has more detailed cost data with two stages of retrofit (standard and advanced) for a sample of 31 dwellings. The formulas used to find the average cost of a retrofit per type of retrofit and archetype are shown in Equations \ref{Eq:RetrofitCost1} and \ref{Eq:RetrofitCost2}.

@@

Where $ {\text{\euro}}{}_{rf-rt,a}$ is the retrofit cost from pre-retrofit BER rating or rating from (rf) to post-retrofit rating or rating to (rt) for archetype (a). $Ali_{r,t}$ is the cost data from  \citet{Ali2020ABuildings}, it contains all BER ratings and retrofit types but does not include an archetype, so an archetype factor (AF) is introduced, 0.9 for apartments, 1 for attached, and 1.1 for detached dwellings.  $TABULA_{rf-rt,a}$ is the average retrofit cost from pre-retrofit BER rating to post-retrofit BER rating for an archetype, data is sourced from TABULA \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland}.  However with only 62 samples in total not all retrofits are included. In this case, TABULA is ignored and \citet{Ali2020ABuildings} is the only source of retrofit costs. The retrofit costs are cross-checked with the residential cost-optimal report \cite{AECOM2020ReportBuildings}. 
\par To simplify the model, the cost of retrofit per BER rating is not used in TIM. Instead, an average weighted cost per archetype and per retrofit is calculated.
